```{=html}
<!-- Note: Section files must have no YAML header and only Report.qmd should have one. 
# If you have trouble compiling, try Create with File → New File → Text File (not Quarto Document, which inserts a header)
# Save with a .qmd file -->
```


The Palmer penguins data set consists of three main species (Adelie, Chinstrap, Gentoo) that reside on 3 main islands (Torgensen,Dream, Biscoe) in the Antartic. This data has been collected by researchers at the Palmer station in Antartica between the years of 2007-2009, this is a United States research station that's situated at the Anvers Island in Antartica. The data contains a mix of categorical (sex, Island and species) and numerical data (Bill length, Flipper length, Body mass etc) (Palmer Station Antarctica LTER & Gorman, 2020), making this data set helpful for statistical analysis as you can pull inferences out of the categorical data using the numerical data such as comparing bill length among species or body mass among sexes. For this data set, there are a limited number of sources, however the sources present show that the palmer data set has shown that this data set has proven very useful when teaching and demonstrating statistical analysis and visualizing data in R, the purpose of this source was to show how this data can be used for various types of data exoploration and anlysis, the authors used the data to make several inferences on all the variables to prove this such as what is being done in this report (Horst e al., 2022). Research has also been done that is similar to the inference we are making in this report such as measuring body characteristics between same species and using those inferences to determine how males and females obtain food, models from this source correctly classified approximately 90 percent of individuals from this data set with several infeernces made about sex between species (Gorman et al., 2014). These sources show that this data set is especially good for this research question as the issue has been tackled before.

The males and females of a similar species often vary in size, males are usually bigger. This would mean that a big male of a very small species could be similar in size to a small female of a bigger species. If we attempt to separate these species by just looking at the sizes of the species and ignore the sexes of the penguins, some penguins might end up getting misclassified. This question aims to check if that actually occurs in the data, and whether this inclusion of sexes in the analysis would help reduce those misclassifications.

# **Descriptive statistics**

As per our question, the first analysis we'll have to make is how results differ between species before incorporating sex, this can be done through the by function that gives us a summary of the data by a categorical variable which in our case will be the species firstly:

```{r}
#This line of R code will give us the summary of the penguins by species to analyse any differences and similarities
penguins_cleaned <- read.csv("penguins_cleaned.csv")
by(penguins_cleaned, penguins_cleaned$species, summary)
```

```{r}

library(ggplot2)
ggplot(penguins_cleaned, 
       aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Body mass (g)"
  )

```

```{r}
ggplot(penguins_cleaned, 
       aes(x = species, y = flipper_length_mm)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Flipper length (mm)"
  )
```

```{r}
ggplot(penguins_cleaned, 
       aes(x = species, y = bill_length_mm)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Bill length (mm)"
  )

```

```{r}
ggplot(penguins_cleaned, 
       aes(x = species, y = bill_depth_mm)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Bill depth (mm)"
  )
```


The first set of statistics show all the descriptive statistics that are related to purely the three species of the penguins, these statistics can be compared against each other to identify which species are most like each other and which aren't to establish base observations before continuing onto analyzing each of the sexes of the species against each other. Starting with the body mass of each three species, when referring to the first graph comparing body mass by species its evident that the two species of Adelie and Chinstrap have the most similar body mass distributions with both species having inter lapping IQRs (of around 3300g to 4000g) with a similar median (between 3700g - 3800g), this cant be said for the Gentoo species who's median is significantly heavier than both other species maximum value at 5000g. The means for Adelie and Chinstrap are also within a 100g of each other while Gentoo sits 2000g above both others at 5090g meaning that Gentoo is significantly larger and heavier than both other species.

These similarities between Adelie and Chinstrap continue when looking at the graph for flipper length vs species as both Adelie and Chinstrap have average length flippers however Chinstrap is slightly larger as the Adelie has a median of 190.1mm while Chinstrap sits at 196mm however we can say that their flipper size distributions are similar since their IQRs (Inter-Quartile Ranges) intersect and are of similar size, the same once again cant be said for Gentoo as they possess larger flippers with their median being a larger 216mm with a IQR being above 200mm. These are the main two attributes when looking at the body size of a penguin however looking at the attributes of bill depth as its graph shows the same pattern of Gentoo being dissimilar to the other two species, the only anomlie in terms of the variables is bill length, this is shown through the graph showcasing the case of bill length vs species, where Adelie has a smaller bill size with the species central tendency being under 45mm with a median of 38.85mm, meanwhile Chinstrap and Gentoo have intersecting IQRs while also being above 45mm with their medians sitting between 45mm-50mm. These results consistently show that the two species of Adelie and Chinstrap being similar in bodily attributes while Gentoo consistently shows to be different, this inference can be used as base knowledge to refer back to when comparing the different sexes of each species.
