
<!--## Methods-->

To answer the question, we broke it into 2 parts: 

1) Do size differences within female and male penguins in one species, cause misclassification for another species; and 

2) Does removing and then adding sex change misclassification?


## Part 1: Sex differences within species

Penguins were removed from the dataset if they had no sex recorded to ensure consistency. A row_id column was also added to help with identifying individual penguins.


```{r}
#| echo: false
#| eval: true
palmer_penguins <- read.csv("penguins_cleaned.csv")

palmer_penguins$row_id <- 1:nrow(palmer_penguins)

#load as factor
palmer_penguins$species <- as.factor(palmer_penguins$species)
palmer_penguins$sex <- as.factor(palmer_penguins$sex)

```


We wanted to first perform a t-test to determine whether there was a significant difference between the mean measurements of each species, accounting for males and females. This approach follows R code demonstrated in a STAT394 lecture (Frery, 2026).

However, to run the t-tests required first performing an F-test to determine whether males and females within each species have equal variance. Results shown below.


```{r}
#| echo: false
#| eval: true
#| results: 'hide'

#Adelie

#bill_length_mm
adelie_m <- subset(palmer_penguins, species=="Adelie" & sex=="male")$bill_length_mm
adelie_f <- subset(palmer_penguins, species=="Adelie" & sex=="female")$bill_length_mm
var.test(adelie_m, adelie_f, alternative = "two.sided")

#p-value = 0.3296


#bill_depth_mm
adelie_m <- subset(palmer_penguins, species=="Adelie" & sex=="male")$bill_depth_mm
adelie_f <- subset(palmer_penguins, species=="Adelie" & sex=="female")$bill_depth_mm
var.test(adelie_m, adelie_f, alternative = "two.sided")

#p-value = 0.513

#flipper_length_mm
adelie_m <- subset(palmer_penguins, species=="Adelie" & sex=="male")$flipper_length_mm
adelie_f <- subset(palmer_penguins, species=="Adelie" & sex=="female")$flipper_length_mm
var.test(adelie_m, adelie_f, alternative = "two.sided")

#p-value = 0.1637

#body_mass_g
adelie_m <- subset(palmer_penguins, species=="Adelie" & sex=="male")$body_mass_g
adelie_f <- subset(palmer_penguins, species=="Adelie" & sex=="female")$body_mass_g
var.test(adelie_m, adelie_f, alternative = "two.sided")

#p-value = 0.03356
 
  
#Gentoo


#bill_length_mm
gentoo_m <- subset(palmer_penguins, species=="Gentoo" & sex=="male")$bill_length_mm
gentoo_f <- subset(palmer_penguins, species=="Gentoo" & sex=="female")$bill_length_mm
var.test(gentoo_m, gentoo_f, alternative = "two.sided")

#p-value = 0.03327

#bill_depth_mm
gentoo_m <- subset(palmer_penguins, species=="Gentoo" & sex=="male")$bill_depth_mm
gentoo_f <- subset(palmer_penguins, species=="Gentoo" & sex=="female")$bill_depth_mm
var.test(gentoo_m, gentoo_f, alternative = "two.sided")

#p-value = 0.0174


#flipper_length_mm
gentoo_m <- subset(palmer_penguins, species=="Gentoo" & sex=="male")$flipper_length_mm
gentoo_f <- subset(palmer_penguins, species=="Gentoo" & sex=="female")$flipper_length_mm

var.test(gentoo_m, gentoo_f, alternative = "two.sided")

#p-value = 0.004889


#body_mass_g
gentoo_m <- subset(palmer_penguins, species=="Gentoo" & sex=="male")$body_mass_g
gentoo_f <- subset(palmer_penguins, species=="Gentoo" & sex=="female")$body_mass_g

var.test(gentoo_m, gentoo_f, alternative = "two.sided")

#p-value = 0.4205

#Chinstrap
  
#bill_length_mm
chinstrap_m <- subset(palmer_penguins, species=="Chinstrap" & sex=="male")$bill_length_mm
chinstrap_f <- subset(palmer_penguins, species=="Chinstrap" & sex=="female")$bill_length_mm

var.test(chinstrap_m, chinstrap_f, alternative = "two.sided")

#p-value = 0.0001593


#bill_depth_mm
chinstrap_m <- subset(palmer_penguins, species=="Chinstrap" & sex=="male")$bill_depth_mm
chinstrap_f <- subset(palmer_penguins, species=="Chinstrap" & sex=="female")$bill_depth_mm

var.test(chinstrap_m, chinstrap_f, alternative = "two.sided")

#p-value = 0.8833

#flipper_length_mm
chinstrap_m <- subset(palmer_penguins, species=="Chinstrap" & sex=="male")$flipper_length_mm
chinstrap_f <- subset(palmer_penguins, species=="Chinstrap" & sex=="female")$flipper_length_mm

var.test(chinstrap_m, chinstrap_f, alternative = "two.sided")

#p-value = 0.8288


#body_mass_g
chinstrap_m <- subset(palmer_penguins, species=="Chinstrap" & sex=="male")$body_mass_g
chinstrap_f <- subset(palmer_penguins, species=="Chinstrap" & sex=="female")$body_mass_g

var.test(chinstrap_m, chinstrap_f, alternative = "two.sided")

# p-value = 0.1762

```

| Species | Measurement | p-value | Equal variance? |
|---|---|---|---|
| Adelie | Bill length | 0.330 | Yes |
| Adelie | Bill depth | 0.513 | Yes |
| Adelie | Flipper length | 0.164 | Yes |
| Adelie | Body mass | 0.034 | No |
| Gentoo | Bill length | 0.033 | No |
| Gentoo | Bill depth | 0.017 | No |
| Gentoo | Flipper length | 0.005 | No |
| Gentoo | Body mass | 0.421 | Yes |
| Chinstrap | Bill length | 0.0002 | No |
| Chinstrap | Bill depth | 0.883 | Yes |
| Chinstrap | Flipper length | 0.829 | Yes |
| Chinstrap | Body mass | 0.176 | Yes |


P values greater than 0.05 indicate where variances are statistically equal.
Based on the R code output above, Adelie body_mass_g, Gentoo bill depth, Gentoo flipper length, Gentoo body mass, Chinstrap bill length all suggest unequal variances. We then set the var.equal parameter in the t-test to FALSE for these measurements.


Next we performed a t-test, following the approach shown by Frery (2026) to check for a significant difference between the means within each species. The t-test divides the difference in sample means by the standard error. Setting equal variances to true or false hence has an effect on the calculation of the standard error and the degrees of freedom used. Results shown below.


```{r}
#| echo: false
#| eval: true
#| results: 'hide'

#Adelie


#Bill_length
t.test(subset(palmer_penguins, species=="Adelie" & sex=="male")$bill_length_mm,
subset(palmer_penguins, species=="Adelie" & sex=="female")$bill_length_mm, var.equal=TRUE)

# p-value = 4.44e-15 smaller than 0.05 suggests that there is a significant difference between males and female bill_length within the species

#bill_depth_mm
t.test(subset(palmer_penguins, species=="Adelie" & sex=="male")$bill_depth_mm,
subset(palmer_penguins, species=="Adelie" & sex=="female")$bill_depth_mm, var.equal=TRUE)

#*p-value = 1.844e-15 suggests significant difference 

#flipper_length_mm
t.test(subset(palmer_penguins, species=="Adelie" & sex=="male")$flipper_length_mm,
subset(palmer_penguins, species=="Adelie" & sex=="female")$flipper_length_mm, var.equal=TRUE)

#p-value = 1.09e-05

#body_mass_g
#UNEQUAL VARIANCES 
t.test(subset(palmer_penguins, species=="Adelie" & sex=="male")$body_mass_g,
subset(palmer_penguins, species=="Adelie" & sex=="female")$body_mass_g, var.equal=FALSE)

#p-value < 2.2e-16

#Gentoo

#Bill_length
t.test(subset(palmer_penguins, species=="Gentoo" & sex=="male")$bill_length_mm,
subset(palmer_penguins, species=="Gentoo" & sex=="female")$bill_length_mm, var.equal=FALSE)

#* p-value = 1.315e-14

#bill_depth_mm
t.test(subset(palmer_penguins, species=="Gentoo" & sex=="male")$bill_depth_mm,
subset(palmer_penguins, species=="Gentoo" & sex=="female")$bill_depth_mm, var.equal=FALSE)

#*p-value < 2.2e-16

#flipper_length_mm
t.test(subset(palmer_penguins, species=="Gentoo" & sex=="male")$flipper_length_mm,
subset(palmer_penguins, species=="Gentoo" & sex=="female")$flipper_length_mm, var.equal=FALSE)

#p-value < 2.2e-16

#body_mass_g
t.test(subset(palmer_penguins, species=="Gentoo" & sex=="male")$body_mass_g,
subset(palmer_penguins, species=="Gentoo" & sex=="female")$body_mass_g, var.equal=TRUE)

#p-value < 2.2e-16



#Chinstrap

#Bill_length
t.test(subset(palmer_penguins, species=="Chinstrap" & sex=="male")$bill_length_mm,
subset(palmer_penguins, species=="Chinstrap" & sex=="female")$bill_length_mm, var.equal=FALSE)

#* p-value = 8.918e-10


#bill_depth_mm
t.test(subset(palmer_penguins, species=="Chinstrap" & sex=="male")$bill_depth_mm,
subset(palmer_penguins, species=="Chinstrap" & sex=="female")$bill_depth_mm, var.equal=TRUE)

#*p-value = 6.564e-13


#flipper_length_mm
t.test(subset(palmer_penguins, species=="Chinstrap" & sex=="male")$flipper_length_mm,
subset(palmer_penguins, species=="Chinstrap" & sex=="female")$flipper_length_mm, var.equal=TRUE)

#p-value = 2.526e-07

#body_mass_g
t.test(subset(palmer_penguins, species=="Chinstrap" & sex=="male")$body_mass_g,
subset(palmer_penguins, species=="Chinstrap" & sex=="female")$body_mass_g, var.equal=TRUE)

#p-value = 2.044e-06


```

| Species | Measurement | Mean (male) | Mean (female) | t-statistic | p-value |
|---|---|---|---|---|---|
| Adelie | Bill length (mm) | 40.39 | 37.26 | 8.78 | 4.44e-15 |
| Adelie | Bill depth (mm) | 19.07 | 17.62 | 8.93 | 1.84e-15 |
| Adelie | Flipper length (mm) | 192.41 | 187.79 | 4.56 | 1.09e-05 |
| Adelie | Body mass (g) | 4043.49 | 3368.84 | 13.13 | < 2.2e-16 |
| Gentoo | Bill length (mm) | 49.47 | 45.56 | 8.88 | 1.32e-14 |
| Gentoo | Bill depth (mm) | 15.72 | 14.24 | 12.49 | < 2.2e-16 |
| Gentoo | Flipper length (mm) | 221.54 | 212.71 | 9.94 | < 2.2e-16 |
| Gentoo | Body mass (g) | 5484.84 | 4679.74 | 14.72 | < 2.2e-16 |
| Chinstrap | Bill length (mm) | 51.09 | 46.57 | 7.57 | 8.92e-10 |
| Chinstrap | Bill depth (mm) | 19.25 | 17.59 | 8.90 | 6.56e-13 |
| Chinstrap | Flipper length (mm) | 199.91 | 191.74 | 5.75 | 2.53e-07 |
| Chinstrap | Body mass (g) | 3938.97 | 3527.21 | 5.21 | 2.04e-06 |

Next we looked at Hotelling's T^2 test, based on R code from the lecture presented by Frery (2026), to examine all the measurements together grouped. Here we compared the mean as vectors, since a penguin's measurements are related - i.e., you would expect a larger penguin to be heavier.


```{r}
#| echo: false
#| eval: true
#| message: false
#| results: 'hide'

# extract measurement columns
library(Hotelling)

#adelie

#grab by column index for measurement columns
adelie_m <- subset(palmer_penguins, species == "Adelie" & sex == "male")[, 3:6]
adelie_f <- subset(palmer_penguins, species == "Adelie" & sex == "female")[, 3:6]
adelie_hotelling <- hotelling.test(adelie_m, adelie_f)
adelie_hotelling


#gentoo
gentoo_m <- subset(palmer_penguins, species == "Gentoo" & sex == "male")[, 3:6]
gentoo_f <- subset(palmer_penguins, species == "Gentoo" & sex == "female")[, 3:6]
gentoo_hotelling <-hotelling.test(gentoo_m, gentoo_f)
gentoo_hotelling



#chinstrap
chinstrap_m <- subset(palmer_penguins, species == "Chinstrap" & sex == "male")[, 3:6]
chinstrap_f <- subset(palmer_penguins, species == "Chinstrap" & sex == "female")[, 3:6]
chinstrap_hotelling <- hotelling.test(chinstrap_m, chinstrap_f)
chinstrap_hotelling



```

| Species | Test statistic | Numerator df | Denominator df | p-value |
|---|---|---|---|---|
| Adelie | 245.09 | 4 | 141 | ≈ 0 |
| Gentoo | 292.15 | 4 | 114 | ≈ 0 |
| Chinstrap | 115.23 | 4 | 63 | 3.19e-13 |


We then ran the Box M test next to determine if the covariance matrices (sigma) across different groups are equal. The test will check spreads and variation
to then determine sigma overall or per species.


```{r}
#| echo: false
#| eval: true
#| message: false
#| results: 'hide'

library(biotools)


adelie <- droplevels(subset(palmer_penguins, species=="Adelie"))
boxM(adelie[, 3:6], adelie$sex) #grab by column index for measurement columns

#p-value = 0.06117

gentoo <- droplevels(subset(palmer_penguins, species=="Gentoo"))
boxM(gentoo[, 3:6], gentoo$sex) #grab by column index for measurement columns

#p-value = 0.003452

chinstrap <- droplevels(subset(palmer_penguins, species=="Chinstrap"))
boxM(chinstrap[, 3:6], chinstrap$sex) #grab by column index for measurement columns

#p-value = 0.00647




```

| Species | Chi-Sq (approx.) | df | p-value | Significant at α=0.05? |
|---|---|---|---|---|
| Adelie | 17.65 | 10 | 0.061 | No (borderline) |
| Gentoo | 26.22 | 10 | 0.0035 | Yes |
| Chinstrap | 24.46 | 10 | 0.0065 | Yes |

An alpha value p=0.05 is used for this test. The null hypothesis is rejected where the p value falls below this threshold, suggesting that males and females have significantly different covariance shapes. We failed to reject the null hypothesis for Adelie but reject for Gentoo and Chinstrap. This suggests that Gentoo and Chinstrap have significant covariance structures.





## Part 2: Classification with sex and without sex
We then progressed with the second part of our question:

First, we calculated the average size of the penguins and spread for each species to find the Mahalanobis distance, based on R code used by Frery (2026).

```{r}
#| echo: false
#| eval: true
#| message: false
#| results: 'hide'


#Adelie
mean.adelie <- colMeans(subset(palmer_penguins, species=="Adelie")[, 3:6])
mean.adelie
covariance.adelie <- var(subset(palmer_penguins, species=="Adelie")[, 3:6])
covariance.adelie

#Gentoo
mean.gentoo   <- colMeans(subset(palmer_penguins, species=="Gentoo")[, 3:6])
mean.gentoo
covariance.gentoo <- var(subset(palmer_penguins, species=="Gentoo")[, 3:6])
covariance.gentoo


#Chinstrap
mean.chinstrap   <- colMeans(subset(palmer_penguins, species=="Chinstrap")[, 3:6])
mean.chinstrap
covariance.chinstrap <- var(subset(palmer_penguins, species=="Chinstrap")[, 3:6])
covariance.chinstrap

```


We used the Mahalanobis distance to find the distance of an individual penguin from the mean vector for the species with relation to the spread. This will show us how similar individuals seem according to each species.

Covariance is also a parameter of this, hence we performed a Box M test again to examine covariances of species to determine whether we used separate or shared covariance matrices.


```{r}
#| echo: false
#| eval: true
#| message: false
#| results: 'hide'

boxM(palmer_penguins[, 3:6], palmer_penguins$species)


```

We initially considered using a chi-squared test but it only compares a penguin to its own species rather than to all the groups available. We changed the method as we could not accurately compare across species.

```{r}
#| echo: true
#| eval: true
#| results: 'hide'

d1.adelie    <- mahalanobis(palmer_penguins[, 3:6], center = mean.adelie,    cov = covariance.adelie)
d2.gentoo    <- mahalanobis(palmer_penguins[, 3:6], center = mean.gentoo,    cov = covariance.gentoo)
d3.chinstrap <- mahalanobis(palmer_penguins[, 3:6], center = mean.chinstrap, cov = covariance.chinstrap)

combined <- cbind(d1.adelie, d2.gentoo, d3.chinstrap)
species_labels <- c("Adelie", "Gentoo", "Chinstrap")
combined_df <- species_labels[apply(combined, 1, which.min)]
combined_df



```

```{r}
#| echo: false
#| eval: true
#| results: 'hide'


misclassified <- which(combined_df != palmer_penguins$species)
length(misclassified)


```
Here each individual penguin is compared on its measurements to the centroids of each species and finds the one with the closest distance (most similar).


We then repeated this model, splitting each species by sex, to see if there would be any change to the misclassified penguins.


```{r}
#| echo: false
#| eval: true
#| results: 'hide'


#adelie

mean.adelie.m    <- colMeans(subset(palmer_penguins, species=="Adelie" & sex=="male")[, 3:6])
covariance.adelie.m <- var(subset(palmer_penguins, species=="Adelie" & sex=="male")[, 3:6])

mean.adelie.f    <- colMeans(subset(palmer_penguins, species=="Adelie" & sex=="female")[, 3:6])
covariance.adelie.f <- var(subset(palmer_penguins, species=="Adelie" & sex=="female")[, 3:6])

d1m.adelie <- mahalanobis(palmer_penguins[, 3:6], center = mean.adelie.m, cov = covariance.adelie.m)
d1f.adelie <- mahalanobis(palmer_penguins[, 3:6], center = mean.adelie.f, cov = covariance.adelie.f)



#Gentoo
mean.gentoo.m    <- colMeans(subset(palmer_penguins, species=="Gentoo" & sex=="male")[, 3:6])
covariance.gentoo.m <- var(subset(palmer_penguins, species=="Gentoo" & sex=="male")[, 3:6])

mean.gentoo.f    <- colMeans(subset(palmer_penguins, species=="Gentoo" & sex=="female")[, 3:6])
covariance.gentoo.f <- var(subset(palmer_penguins, species=="Gentoo" & sex=="female")[, 3:6])

d2m.gentoo <- mahalanobis(palmer_penguins[, 3:6], center = mean.gentoo.m, cov = covariance.gentoo.m)
d2f.gentoo <- mahalanobis(palmer_penguins[, 3:6], center = mean.gentoo.f, cov = covariance.gentoo.f)


#Chinstrap

mean.chinstrap.m    <- colMeans(subset(palmer_penguins, species=="Chinstrap" & sex=="male")[, 3:6])
covariance.chinstrap.m <- var(subset(palmer_penguins, species=="Chinstrap" & sex=="male")[, 3:6])

mean.chinstrap.f    <- colMeans(subset(palmer_penguins, species=="Chinstrap" & sex=="female")[, 3:6])
covariance.chinstrap.f <- var(subset(palmer_penguins, species=="Chinstrap" & sex=="female")[, 3:6])

d3m.chinstrap <- mahalanobis(palmer_penguins[, 3:6], center = mean.chinstrap.m, cov = covariance.chinstrap.m)
d3f.chinstrap <- mahalanobis(palmer_penguins[, 3:6], center = mean.chinstrap.f, cov = covariance.chinstrap.f)


combined_sex <- cbind(d1m.adelie, d1f.adelie, d2m.gentoo, d2f.gentoo, d3m.chinstrap, d3f.chinstrap)

group_labels <- c("Adelie","Adelie","Gentoo","Gentoo","Chinstrap","Chinstrap")

combined_sex_df <- group_labels[apply(combined_sex, 1, which.min)]


misclassified_sex <- which(combined_sex_df != palmer_penguins$species)
length(misclassified_sex)



misclassified_df_final <- data.frame(
  row_id = 1:nrow(palmer_penguins),
  species = palmer_penguins$species,
  sex = palmer_penguins$sex,
  predicted = combined_df,
  correct = combined_df == palmer_penguins$species
)

misclassified_sex_df_final <- data.frame(
  row_id = 1:nrow(palmer_penguins),
  species = palmer_penguins$species,
  sex = palmer_penguins$sex,
  predicted = combined_sex_df,
  correct = combined_sex_df == palmer_penguins$species
)


#misclassified penguins with and without sex
#head(misclassified_df_final)
#head(misclassified_sex_df_final)


#filtered
filtered_misclassified_df_final <- subset(misclassified_df_final, correct == FALSE)
filtered_misclassified_sex_df_final   <- subset(misclassified_sex_df_final, correct == FALSE)

```

```{r}
#| echo: true
#| eval: true
#| results: 'hide'


#adelie
d1m.adelie <- mahalanobis(palmer_penguins[, 3:6], center = mean.adelie.m, cov = covariance.adelie.m)
d1f.adelie <- mahalanobis(palmer_penguins[, 3:6], center = mean.adelie.f, cov = covariance.adelie.f)



#Gentoo
d2m.gentoo <- mahalanobis(palmer_penguins[, 3:6], center = mean.gentoo.m, cov = covariance.gentoo.m)
d2f.gentoo <- mahalanobis(palmer_penguins[, 3:6], center = mean.gentoo.f, cov = covariance.gentoo.f)


#Chinstrap
d3m.chinstrap <- mahalanobis(palmer_penguins[, 3:6], center = mean.chinstrap.m, cov = covariance.chinstrap.m)
d3f.chinstrap <- mahalanobis(palmer_penguins[, 3:6], center = mean.chinstrap.f, cov = covariance.chinstrap.f)


combined_sex <- cbind(d1m.adelie, d1f.adelie, d2m.gentoo, d2f.gentoo, d3m.chinstrap, d3f.chinstrap)

group_labels <- c("Adelie","Adelie","Gentoo","Gentoo","Chinstrap","Chinstrap")

combined_sex_df <- group_labels[apply(combined_sex, 1, which.min)]



```


```{r}
#| echo: false
#| eval: true
#| results: 'hide'


# check misclassified penguins from the loaded data
check_misclassified_4 <- subset(palmer_penguins, row_id %in% c(68, 124, 286, 296))
check_misclassified_4
check_misclassified_1 <- subset(palmer_penguins, row_id %in% 274)
check_misclassified_1


# generate mean values again for examination
#mean.adelie.f
#mean.chinstrap.f
#mean.adelie.m
#mean.chinstrap.m

#mean.adelie
#mean.chinstrap


```


